smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies

### 2023-09-06 manuscript/manuscript.qmd
implemented FT comments

[**Word count:**]{.underline} 1,970 words/2,000 words

@@

145 words/150 words

@@

Partially observed confounder data are a major challenge for the statistical analysis of electronic health records (EHR). While analytic approaches exist (e.g., multiple imputation), assumptions on underlying missingness patterns and mechanisms must hold. We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.

@@

The `smdi` R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can give valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.

@@

As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative purposes (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records). Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are usually not available for all patients at all necessary time points which leads to missing data. If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, this can lead to bias in the resulting effect estimates for the studied treatments if common missing data analysis methods like complete case analysis or imputation are used. Hence, it is of utmost importance to investigate the potential patterns and mechanisms to know if assumptions for common analytic methods hold. Here, we present the `smdi` R package, which enables researchers to perform such principled missing data investigations on partially observed confounders. The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on observed data.
:::

@@

Electronic health records (EHR) are increasingly used to conduct real-world evidence (RWE) studies to complement evidence from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their detailed capture of clinical parameters, EHRs can significantly improve the ability to control for imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often partially observed which challenges the statistical analysis and can result in severe bias when estimating treatment effects if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study, but guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

Given these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data.

@@

To streamline implementation of routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].

@@

Missing data are ubiquitous in EHR and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data.

The package also has limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data. Hence, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.

@@

The smdi package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies. This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.

>>>>

[**Word count:**]{.underline} 1,983 words/2,000 words

@@

150 words/150 words

@@

Partially observed confounder data pose a major challenge when conducting causal inference and statistical analyses using electronic health records (EHR). While analytic approaches such as multiple imputation are available, assumptions on underlying missingness patterns and mechanisms must hold. We aimed to develop a toolkit to streamline missing data diagnostics to scrutinize if certain analytic approaches are viable options.

@@

The `smdi` R package is freely available on CRAN and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can provide valuable insights into underlying missingness patterns and mechanisms and thereby help improve the robustness of real-world evidence studies.

@@

As opposed to clinical trials, which are designed to collect data for research purposes in a harmonized manner, real-world data are typically generated for administrative (e.g., health insurance claims for billing purposes) or clinical documentation (e.g., electronic health records) purposes. Hence, confounders and prognostic factors, which need to be balanced between treatment groups to infer causal treatment effects, are often unavailable for all patients at all necessary time points, resulting in missing data. If the underlying missingness mechanisms of such factors is not at random, e.g., patients with higher levels of a prognostic biomarker are more likely to have missing data, it can introduce bias into the resulting effect estimates for the studied treatments when common missing data analysis methods like complete case analysis or imputation are applied. Therefore, it is of utmost importance to investigate potential missingness patterns and mechanisms to determine if assumptions for common analytic methods hold. Here, we present the `smdi` R package, which enables researchers to conduct such principled missing data investigations on partially observed confounders. The `smdi` package implements functions to visualize and describe missing data and to infer potential missingness patterns and mechanisms based on observed data.
:::

@@

Electronic health records (EHR) are increasingly used to generate real-world evidence (RWE) to complement data from randomized controlled trials (RCTs) [@fdaRWE2018; @Desai2021]. Due to their comprehensive capture of clinical parameters, EHRs can significantly improve the ability to mitigate imbalances in prognostic factors between treatment groups, especially when linked to administrative claims databases [@asfaw2022]. However, such prognostic factors are often partially observed, posing a challenge to statistical analysis and potentially leading to severe bias in treatment effect estimates if not handled appropriately [@gorelick2006; @ayilara2019; @groenwold2020].

In order to make informed decisions about the most appropriate analytic approach, it is crucial to investigate the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study, but guidance frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] using directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far only been described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently infrequently performed [@carroll2020; @wood2004; @harel2012].

@@

Considering these limitations, we have recently developed and evaluated a principled approach combining multiple missing data diagnostics using an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics effectively characterized missing data mechanisms and provided helpful guidance for the choice of appropriate analytic methods to handle POC data.

@@

To streamline the implementation of routine missing data diagnostics for POC data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi].

@@

Missing data are ubiquitous in RWE studies involving EHR and may introduce bias if not handled appropriately. To address this issue, we developed the `smdi` R package to streamline routine diagnostics of missing data.

The package does come with certain limitations, such that the true underlying mechanism causing the missing data can never be inferred with absolute certainty from the observed data. Therefore, it's important to complement diagnostic results with substantive expert knowledge to factor in how covariates are measured in routine care and contextualize potential reasons for missingness.

@@

The smdi R package is a powerful and convenient tool to implement principled routine missing data diagnostics in RWE studies. This will improve the robustness of studies involving POCs by elucidating if certain analytic approaches are viable for a given dataset.
